---
title: "Getting Started with usdeaths"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with usdeaths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

The `usdeaths` package provides tools for downloading, reading, and decoding
CDC vital statistics mortality files. These are fixed-width files published by
the National Center for Health Statistics (NCHS) and cover US death records
going back to the 1960s. Because the raw files use numeric codes rather than
human-readable labels, working with them directly is cumbersome. This package
handles the download, parsing, and decoding for you.

## Setup

```{r setup, message=FALSE}
library(usdeaths)
```

## Quick Start

For most users, `cdc_import()` is all you need. Pass it a section name and
year and it handles everything — resolving the download URL, fetching the file
from the CDC FTP server, reading the fixed-width format, and decoding all
coded columns into human-readable labels:

```{r import, eval=FALSE}
mort1969 <- cdc_import("mortality_multiple", 1969)
#> Decoding 1921990 rows...
#> # A tibble: 1,921,990 × 43
#>    last_digit_data_year shipment_number reporting_area  certificate_number ...
#>    <chr>                <chr>           <chr>           <chr>
#>  1 1969                 13              All other areas NA
#>  2 1969                 05              All other areas NA
#>  ...
```

Note: these files can be several hundred megabytes and contain millions of
records. The 1969 mortality file has nearly 2 million rows. Download and
decode time will vary depending on your connection and hardware.

## Under the Hood

`cdc_import()` is a wrapper around four lower-level functions. You can call
these individually if you need more control over any step.

### Step 1: Resolve the URL

`get_cdc_url()` looks up the download URL for a given section and year from
the internal CDC link table:

```{r url, eval=FALSE}
url <- get_cdc_url("mortality_multiple", 1969)
```

### Step 2: Download

`download_cdc()` fetches the zip file from the CDC FTP server and returns a
path to a temporary file:

```{r download, eval=FALSE}
temp <- download_cdc(url)
```

### Step 3: Read

`load_data()` needs the metadata for the year you are working with. Each
year's file has its own fixed-width layout and code dictionary, shipped as
a data object in the package:

```{r meta, eval=FALSE}
meta <- data_mortality_multiple_1969
mort1969 <- load_data(temp, meta)
```

The raw data comes back with numeric codes in every column:

```{r raw, eval=FALSE}
mort1969
#> # A tibble: 1,921,990 × 43
#>    last_digit_data_year shipment_number reporting_area certificate_number ...
#>    <chr>                <chr>           <chr>          <chr>
#>  1 9                    13              0              NA
#>  2 9                    05              0              NA
#>  ...
```

### Step 4: Decode

`decode_all()` decodes every coded column in place and returns the full
dataset with human-readable labels:

```{r decode, eval=FALSE}
mort1969_decoded <- decode_all(mort1969, meta)
```

## Inspecting Before Decoding

For large files you may want to verify the codes are mapping correctly before
committing to a full decode. `decode_preview()` lets you inspect a subset of
columns and rows side by side with their decoded labels:

```{r preview-default, eval=FALSE}
# first 5 coded columns, 1000 rows
decode_preview(mort1969, meta)

# first 10 coded columns
decode_preview(mort1969, meta, first_n = 10)

# first 3 coded columns plus columns 12 and 20 specifically
decode_preview(mort1969, meta, first_n = 3, numbers = c(12, 20))
```
